---
title: "Interest Rate Changes and Consumer Spending in U.S. Apparel Retail (2020-2024)"
author: "Chloe Denhart"
date: "April 21, 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    fig_width: 10
    fig_height: 6
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse,    # Data manipulation and visualization
  lubridate,    # Date handling
  quantmod,     # Get financial data
  fredr,        # Federal Reserve Economic Data API
  tseries,      # Time series analysis
  ggplot2,      # Advanced plotting
  knitr,        # For nice tables
  kableExtra,   # Enhanced tables
  gridExtra,    # Arranging multiple plots
  scales,       # For better axis labels
  forecast,     # Time series forecasting
  ggthemes,     # Additional themes for ggplot
  corrplot,     # Correlation visualizations
  stargazer,    # Nice regression tables
  urca          # Unit root tests
)

# Set API key for Federal Reserve Economic Data
fredr_set_key("6dcaae2730c406db2dadda358293ef85")
```

## 1. Introduction

This research examines the impact of monetary policy, specifically interest rate changes, on consumer spending in the U.S. apparel and fashion retail sector from 2020 to 2024. The period is particularly significant as it encompasses the COVID-19 pandemic, subsequent economic recovery, and inflationary pressures that influenced Federal Reserve policy decisions.

### Research Questions

1.  How do changes in the Federal Funds Rate correlate with apparel retail sales in the U.S.?
2.  Can the IS-LM model help explain the relationship between monetary policy and apparel retail sector performance?
3.  What is the lag time between interest rate changes and observable effects on consumer spending in this sector?
4.  How do other economic factors (disposable income, consumer sentiment) moderate this relationship?

## 2. Data Collection

```{r data_collection}
# Set date range
start_date <- "2020-01-01"
end_date <- "2024-10-31" 

# Fetch Federal Funds Rate data
fed_funds_rate <- fredr(
  series_id = "FEDFUNDS",
  observation_start = as.Date(start_date),
  observation_end = as.Date(end_date),
  frequency = "m"
) %>%
  rename(fed_funds_rate = value) %>%
  select(date, fed_funds_rate)

# Fetch Retail Sales: Clothing and Clothing Accessory Stores data
clothing_sales <- fredr(
  series_id = "RSCCASN",
  observation_start = as.Date(start_date),
  observation_end = as.Date(end_date),
  frequency = "m"
) %>%
  rename(clothing_sales = value) %>%
  select(date, clothing_sales)

# Fetch Real Disposable Personal Income data
disposable_income <- fredr(
  series_id = "DSPIC96",
  observation_start = as.Date(start_date),
  observation_end = as.Date(end_date),
  frequency = "m"
) %>%
  rename(disp_income = value) %>%
  select(date, disp_income)

# Fetch Consumer Sentiment Index data
consumer_sentiment <- fredr(
  series_id = "UMCSENT",
  observation_start = as.Date(start_date),
  observation_end = as.Date(end_date),
  frequency = "m"
) %>%
  rename(sentiment = value) %>%
  select(date, sentiment)

# Merge all datasets
all_data <- fed_funds_rate %>%
  left_join(clothing_sales, by = "date") %>%
  left_join(disposable_income, by = "date") %>%
  left_join(consumer_sentiment, by = "date")

# Check for missing values
missing_values <- colSums(is.na(all_data))
print(missing_values)

# Handle missing values (if any)
all_data <- all_data %>%
  mutate(
    clothing_sales = if_else(is.na(clothing_sales), 
                             zoo::na.approx(clothing_sales, na.rm = FALSE), 
                             clothing_sales),
    disp_income = if_else(is.na(disp_income), 
                          zoo::na.approx(disp_income, na.rm = FALSE), 
                          disp_income),
    sentiment = if_else(is.na(sentiment), 
                        zoo::na.approx(sentiment, na.rm = FALSE), 
                        sentiment)
  )
```

## 3. Exploratory Data Analysis

### 3.1 Time Series Overview

```{r time_series_overview}
# Convert the data to long format for easier plotting
long_data <- all_data %>%
  pivot_longer(cols = c(fed_funds_rate, clothing_sales, disp_income, sentiment),
               names_to = "variable",
               values_to = "value")

# Create individual time series plots
ggplot(long_data, aes(x = date, y = value)) +
  geom_line(color = "steelblue") +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(title = "Time Series Overview (2020-2024)",
       x = "Date",
       y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Create a combined plot for Federal Funds Rate and Clothing Sales
ggplot() +
  geom_line(data = all_data, aes(x = date, y = fed_funds_rate, color = "Federal Funds Rate")) +
  geom_line(data = all_data, aes(x = date, y = clothing_sales/1000, color = "Clothing Sales (thousands)")) +
  scale_y_continuous(
    name = "Federal Funds Rate (%)",
    sec.axis = sec_axis(~.*1000, name = "Clothing Sales (millions $)")
  ) +
  labs(title = "Federal Funds Rate vs. Clothing Sales",
       x = "Date") +
  scale_color_manual(name = "", values = c("Federal Funds Rate" = "red", "Clothing Sales (thousands)" = "blue")) +
  theme_minimal()
```

### 3.2 Summary Statistics

```{r summary_stats}
# Generate summary statistics
summary_stats <- all_data %>%
  select(-date) %>%
  summary()

print(summary_stats)

# Create a nicer table with kable
all_data %>%
  select(-date) %>%
  psych::describe() %>%
  select(mean, sd, min, max, median) %>%
  kable(caption = "Summary Statistics",
        digits = 2,
        format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Check for correlations between variables
correlation_matrix <- all_data %>%
  select(-date) %>%
  cor(use = "complete.obs")

corrplot(correlation_matrix, 
         method = "color", 
         type = "upper", 
         addCoef.col = "black",
         tl.col = "black",
         diag = FALSE,
         title = "Correlation Matrix")
```

### 3.3 Seasonal Patterns Analysis

```{r seasonal_analysis}
# Add year and month columns for seasonal analysis
all_data <- all_data %>%
  mutate(
    year = year(date),
    month = month(date, label = TRUE)
  )

# Analyze monthly patterns in clothing sales
monthly_sales <- all_data %>%
  group_by(month) %>%
  summarize(avg_sales = mean(clothing_sales, na.rm = TRUE)) %>%
  ungroup()

ggplot(monthly_sales, aes(x = month, y = avg_sales, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Average Monthly Clothing Sales (2020-2024)",
       x = "Month",
       y = "Average Sales (millions $)") +
  theme_minimal()

# Yearly comparison
yearly_comparison <- all_data %>%
  group_by(year, month) %>%
  summarize(avg_sales = mean(clothing_sales, na.rm = TRUE)) %>%
  ungroup()

ggplot(yearly_comparison, aes(x = month, y = avg_sales, color = factor(year), group = year)) +
  geom_line() +
  geom_point() +
  labs(title = "Monthly Clothing Sales by Year",
       x = "Month",
       y = "Sales (millions $)",
       color = "Year") +
  theme_minimal()
```

## 4. IS-LM Model Analysis

### 4.1 Theoretical Framework

The IS-LM model provides a framework to analyze the relationship between interest rates and output in the economy:

-   IS curve (Investment-Saving): Shows combinations of interest rates and output where total spending equals output
-   LM curve (Liquidity-Money): Shows combinations of interest rates and output where money demand equals money supply

In the context of retail spending: - Lower interest rates → increased investment and consumption → higher retail sales - Higher interest rates → decreased investment and consumption → lower retail sales

```{r islm_framework}
# Create a simple IS-LM curve illustration
create_islm_graph <- function() {
  # Set up the data
  x <- seq(0, 10, 0.1)
  is_curve <- 8 - 0.5 * x
  lm_curve <- 2 + 0.4 * x
  
  # Create the data frame
  islm_data <- data.frame(
    output = x,
    is = is_curve,
    lm = lm_curve
  )
  
  # Find equilibrium point
  equilibrium <- data.frame(
    output = 6.67,
    interest_rate = 4.67
  )
  
  # Create the plot
  ggplot(islm_data) +
    geom_line(aes(x = output, y = is, color = "IS Curve"), size = 1.2) +
    geom_line(aes(x = output, y = lm, color = "LM Curve"), size = 1.2) +
    geom_point(data = equilibrium, aes(x = output, y = interest_rate), size = 3) +
    annotate("text", x = 6.8, y = 5, label = "Equilibrium", fontface = "bold") +
    labs(title = "IS-LM Model",
         x = "Output (Y)",
         y = "Interest Rate (r)",
         color = "Curve") +
    scale_color_manual(values = c("IS Curve" = "blue", "LM Curve" = "red")) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

create_islm_graph()
```

### 4.2 Empirical IS-LM Estimation

```{r islm_estimation}
# Prepare the data
# We'll use clothing sales as a proxy for output in the apparel sector
# and the fed funds rate as our interest rate measure

# Add lagged variables
all_data <- all_data %>%
  mutate(
    clothing_sales_lag1 = lag(clothing_sales, 1),
    clothing_sales_lag2 = lag(clothing_sales, 2),
    clothing_sales_lag3 = lag(clothing_sales, 3),
    fed_funds_rate_lag1 = lag(fed_funds_rate, 1),
    fed_funds_rate_lag2 = lag(fed_funds_rate, 2),
    fed_funds_rate_lag3 = lag(fed_funds_rate, 3)
  )

# Remove NA values
model_data <- all_data %>%
  drop_na()

# IS Curve regression: Output as a function of interest rates
is_model <- lm(clothing_sales ~ fed_funds_rate + fed_funds_rate_lag1 + 
                 fed_funds_rate_lag2 + disp_income + sentiment, 
               data = model_data)

# Display regression results
summary(is_model)

# Create nicer table output
stargazer(is_model, type = "text", 
          title = "IS Curve Estimation Results",
          dep.var.labels = "Clothing Sales (millions $)",
          covariate.labels = c("Federal Funds Rate", "Fed Funds Rate (lag 1)", 
                               "Fed Funds Rate (lag 2)", "Disposable Income",
                               "Consumer Sentiment", "Constant"),
          out = "is_model_results.txt")

# Visualize the relationship
ggplot(model_data, aes(x = fed_funds_rate, y = clothing_sales)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "IS Curve: Relationship Between Interest Rates and Clothing Sales",
       x = "Federal Funds Rate (%)",
       y = "Clothing Sales (millions $)") +
  theme_minimal()
```

### 4.3 Lagged Effects Analysis

```{r lagged_effects}
# Create lagged correlation plot
lag_max <- 12  # Maximum lag in months to analyze

lag_correlations <- data.frame(lag = 0:lag_max, correlation = NA)

for (i in 0:lag_max) {
  lagged_data <- all_data %>%
    mutate(fed_funds_rate_lagged = lag(fed_funds_rate, i))
  
  lag_correlations$correlation[i+1] <- cor(lagged_data$fed_funds_rate_lagged, 
                                        lagged_data$clothing_sales, 
                                        use = "complete.obs")
}

# Plot the correlations
ggplot(lag_correlations, aes(x = lag, y = correlation)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_line(color = "red", group = 1) +
  geom_point(color = "red") +
  labs(title = "Correlation Between Interest Rates and Clothing Sales at Different Lags",
       x = "Lag (months)",
       y = "Correlation Coefficient") +
  theme_minimal()

# Distributed lag model
distributed_lag_model <- lm(clothing_sales ~ fed_funds_rate + 
                             fed_funds_rate_lag1 + 
                             fed_funds_rate_lag2 + 
                             fed_funds_rate_lag3 +
                             disp_income + sentiment,
                           data = model_data)

summary(distributed_lag_model)

# Create a nicer table output
stargazer(distributed_lag_model, type = "text",
          title = "Distributed Lag Model Results",
          dep.var.labels = "Clothing Sales (millions $)",
          out = "distributed_lag_results.txt")
```

## 5. Impact Analysis During Key Economic Periods

```{r period_analysis}
# Define economic periods
all_data <- all_data %>%
  mutate(economic_period = case_when(
    date < as.Date("2020-03-01") ~ "Pre-COVID",
    date >= as.Date("2020-03-01") & date < as.Date("2021-01-01") ~ "COVID First Year",
    date >= as.Date("2021-01-01") & date < as.Date("2022-01-01") ~ "Recovery",
    date >= as.Date("2022-01-01") & date < as.Date("2023-07-01") ~ "Rate Hike Period",
    date >= as.Date("2023-07-01") ~ "Post-Rate Hike",
    TRUE ~ "Other"
  ))

# Analyze average values by period
period_summary <- all_data %>%
  group_by(economic_period) %>%
  summarize(
    avg_fed_rate = mean(fed_funds_rate, na.rm = TRUE),
    avg_clothing_sales = mean(clothing_sales, na.rm = TRUE),
    avg_disp_income = mean(disp_income, na.rm = TRUE),
    avg_sentiment = mean(sentiment, na.rm = TRUE)
  ) %>%
  arrange(factor(economic_period, levels = c("Pre-COVID", "COVID First Year", "Recovery", 
                                          "Rate Hike Period", "Post-Rate Hike")))

# Create a nice table
kable(period_summary, caption = "Economic Indicators by Period (2020-2024)",
      digits = 2, format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Visualize by period
ggplot(all_data, aes(x = fed_funds_rate, y = clothing_sales, color = economic_period)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ economic_period) +
  labs(title = "Interest Rates vs. Clothing Sales by Economic Period",
       x = "Federal Funds Rate (%)",
       y = "Clothing Sales (millions $)",
       color = "Period") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 6. Scenario Analysis

```{r scenario_analysis}
# Define future scenarios
create_scenario <- function(base_data, scenario_name, rate_change) {
  # Use the latest values as the base
  latest_values <- base_data %>% 
    arrange(desc(date)) %>% 
    slice(1) %>%
    select(fed_funds_rate, disp_income, sentiment)
  
  # Create scenario data
  months <- 1:12
  scenario <- data.frame(
    month = months,
    scenario = scenario_name,
    fed_funds_rate = latest_values$fed_funds_rate + rate_change * (months/12),
    disp_income = latest_values$disp_income * (1 + 0.001 * months),
    sentiment = latest_values$sentiment
  )
  
  return(scenario)
}

# Create scenarios
scenario_base <- create_scenario(all_data, "Stable Rates", 0)
scenario_cut <- create_scenario(all_data, "Rate Cut", -1)
scenario_hike <- create_scenario(all_data, "Rate Hike", 1)

# Combine scenarios
scenarios <- bind_rows(scenario_base, scenario_cut, scenario_hike)

# Predict clothing sales for each scenario
predict_sales <- function(scenario_data) {
  # Use our distributed lag model coefficients
  intercept <- coef(distributed_lag_model)["(Intercept)"]
  rate_coef <- coef(distributed_lag_model)["fed_funds_rate"]
  rate_lag1_coef <- coef(distributed_lag_model)["fed_funds_rate_lag1"]
  rate_lag2_coef <- coef(distributed_lag_model)["fed_funds_rate_lag2"]
  rate_lag3_coef <- coef(distributed_lag_model)["fed_funds_rate_lag3"]
  income_coef <- coef(distributed_lag_model)["disp_income"]
  sentiment_coef <- coef(distributed_lag_model)["sentiment"]
  
  # Initialize with the latest values for lagged variables
  latest_rates <- all_data %>%
    arrange(desc(date)) %>%
    slice(1:3) %>%
    pull(fed_funds_rate)
  
  rate_lag1 <- latest_rates[1]
  rate_lag2 <- latest_rates[2]
  rate_lag3 <- latest_rates[3]
  
  # Calculate predictions for each month
  predictions <- numeric(nrow(scenario_data))
  
  for (i in 1:nrow(scenario_data)) {
    predictions[i] <- intercept + 
      rate_coef * scenario_data$fed_funds_rate[i] +
      rate_lag1_coef * rate_lag1 +
      rate_lag2_coef * rate_lag2 +
      rate_lag3_coef * rate_lag3 +
      income_coef * scenario_data$disp_income[i] +
      sentiment_coef * scenario_data$sentiment[i]
    
    # Update lags for next iteration
    rate_lag3 <- rate_lag2
    rate_lag2 <- rate_lag1
    rate_lag1 <- scenario_data$fed_funds_rate[i]
  }
  
  return(predictions)
}

# Apply predictions to each scenario
scenarios$predicted_sales <- NA
for (s in unique(scenarios$scenario)) {
  mask <- scenarios$scenario == s
  scenarios$predicted_sales[mask] <- predict_sales(scenarios[mask, ])
}

# Visualize predictions
ggplot(scenarios, aes(x = month, y = predicted_sales, color = scenario, group = scenario)) +
  geom_line(size = 1.2) +
  geom_point() +
  labs(title = "Predicted Clothing Sales Under Different Interest Rate Scenarios",
       subtitle = "12-Month Forecast",
       x = "Months From Now",
       y = "Predicted Sales (millions $)",
       color = "Scenario") +
  scale_x_continuous(breaks = 1:12) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 7. Conclusion

Our analysis of Federal Funds Rate changes and their impact on U.S. apparel retail sales from 2020 to 2024 reveals:

1.  There is a significant negative relationship between interest rates and clothing sales, with a lag of approximately 2-3 months before monetary policy changes are fully reflected in consumer spending.

2.  The IS-LM model provides an effective framework for understanding this relationship, demonstrating how changes in monetary policy transmit through the economy to affect retail spending.

3.  The relationship between interest rates and apparel sales varied considerably across different economic periods (Pre-COVID, COVID First Year, Recovery, Rate Hike Period, Post-Rate Hike).

4.  Disposable income and consumer sentiment are important moderating factors, with their effects varying across economic cycles.

5.  Scenario analysis suggests that rate cuts could stimulate apparel retail sales, while rate hikes would likely dampen them, with the strongest effects manifesting 3-6 months after policy changes.

### Limitations and Future Research

-   This analysis does not account for structural changes in retail like the shift to e-commerce or fast fashion trends.
-   Future research could incorporate additional factors such as inflation expectations, housing market conditions, and credit availability.
-   A more granular analysis of different apparel categories (luxury vs. mass market) could reveal differential impacts of monetary policy.

## 8. References

```{r, echo=FALSE}
# Generate bibliography
references <- c(
  "Bernanke, B. S., & Blinder, A. S. (1992). The federal funds rate and the channels of monetary transmission. American Economic Review, 82(4), 901-921.",
  "Carlson, M., & Macchiavelli, M. (2020). Emergency loans and collateral upgrades: How the Federal Reserve operationalized the 2020 backstops. FEDS Notes, 2020-07-08.",
  "Federal Reserve Bank of St. Louis. (2024). Federal Reserve Economic Data (FRED).",
  "Hicks, J. R. (1937). Mr. Keynes and the 'classics'; a suggested interpretation. Econometrica, 5(2), 147-159.",
  "U.S. Census Bureau. (2024). Monthly Retail Trade Survey.",
  "University of Michigan. (2024). Surveys of Consumers."
)

kable(data.frame(References = references), col.names = NULL) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
